### 2019-02-21 Datelife_paper1.Rmd
some figures

---
output: pdf_document
fontsize: 12pts
textalign: left  # this does not work...
geometry: margin = 1in
bibliography: library_red.bib
csl: systematic-biology.csl
link-citations: yes
# indent: true # this works
# header-includes:
#    - \setlength\parindent{24pt}  # this works too, but I'mg onna indent manually with 4 spaces
# nope, 4 spaces indents are not recognized, they do something else
linestretch: 2
header-includes:
    - \usepackage[left]{lineno}
    - \linenumbers
---
```{r eval = TRUE, echo = FALSE}
fig_workflow_main <- "Fig. 1"
fig_workflow_all_functions <- "Vignette Fig. 1"
fig_runtime_main <- "Fig. 2"  
# make_datelife_query_plus_get_datelife_result
```

@@

<!--INTRODUCTION-->
Time of lineage divergence<!--alt lineages: named biological entities, taxon ages?-->
constitutes in many ways the fundamental/main knowledge necessary for evolutionary
understanding.
<!--Along with phylogenetic relationships, it is the most important piece of information
allowing to test alternative evolutionary hypothesis in many areas of research,
from developmental to ecological questions.
Along with migration rates and patterns, it is fundamental to test biogeographical
hypothesis-->
Coupled to species number and distribution, it is the main information necessary
for the study of diversification processes (i.e., the tempo and mode of speciation
and extinction),
<!-- crucial/vital/critical/essential  -->central for the understanding of how biodiversity
patterns are shaped across space, time and clades [@Morlon2014].
Evolutionary understanding also relies on comparative studies, for which knowing
the time context for all life is crucial. Efforts to have a whole tree of life have
been great and here are some examples.
<!--INTRODUCTION: 2nd paragraph-->
When organisms are preserved in a fossil form, a time frame of taxon origin can
be obtained directly from the age of rock strata.
However, not all organisms fossilize well or at all.
Fossilization success alone is highly circumstancial, and varies depending on a
number of parameters including the nature of the habitat, population size, species
range breadth and physical characteristics of the organism, all of which greatly
vary across different organisms and through time.
<!--fossil recovery is also highly circumstancial-->
Thus, relying only on the fossil record to obtain a time frame of lineage divergence
<!--taxon ages--> for all life is<!--, alas,--> not possible.

Another caveat is that we usually cannot have a point age estimates of lineage origin
because of the nature of the fossilization process: when a fossil first appears
on the fossil record it is not necessarily because the lineage just originated,
it might just be bc the conditions for fossilization started at that point but the
organism might have been around for a long time before that. Because of this, we
use ranges from the age of strata where fossils are found. But this does not solve
the problem. The organism could even have been around previously, but not fossilize
under the conditions prevailing around the time of formation of rock strata below
its first appearance in the fossil record. In this sense, fossil ages can only be
considered as minimum time of origin of lineages.
<!--D Bapst work?-->

Another source that has been widely used to inform about the timing of lineage origin
is the relative rate of DNA or aminoacid substitution.
It is estimated from hypothesis of character homology (alignments) for reconstructing
phylogenetic relationships.
<!--describe the two components of rel subst rate? bc they contain the information
on absolute rate of substitution and absolute time.-->
Molecular dating techniques use external data such as absolute time calibrations
(e.g., fossils, geologic events) or absolute substitution rates <!--obtained in
the lab?? birds-->to <!--dissociate rel subs rates components and--> generate dated
phylogenies (chronograms) which contain information on absolute times of node divergence
and taxon ages. <!--With molecular dating methods, an external calibration point
is still necessary to estimate the absolute time of origin. Hence, the accumulation
of data on fossils for calibration is also continual-->
<!--This method has been particularly useful to obtain time frame of origin of organisms
with poor or no fossil record, but is also largely used for organisms with rich
fossil records-->
<!--Models allow to take into account the nature of the fossil record: Given the
characteristics of the fossilization process, the quality of dates inferred directly
from it is not always considered the best.-->
<!--INTRODUCTION: 3rd paragraph-->
In the past two decades, the possibility to obtain good quality DNA sequences
<!-- from a wide variety of organisms became a reality, which,  -->
coupled to methodological developments in phylogenetic and dating inference, allowed
the application of molecular dating methods on a very large amount and diversity
of organisms, greatly increasing the quantity of<!--organisms with--> data on taxon
ages across the tree of life.
<!-- figure showing growth on studies with age data from references table from ttol-->
To date, there is a large amount of both fossil and molecular-based data on taxon
ages and phylogenetic relationships in public repositories such as Dryad, TreeBASE
and Open Tree of Life (OToL).
OToL alone holds more than 200 chronograms. <!--cache updated jan 15 2018
devtools::load_all(pkg = "~/Desktop/datelife")
length(opentree_chronograms$trees)
names(opentree_chronograms)
-->
<!--How many trees? How many dated trees?-->
Methods to include living and fossil lineages are in continued development and increased
usage by the community<!--total evidence and revbayes (fossilized birth-death)-->,
which coupled to better sharing data practices, are greatly contributing to the
accumulation in number and type of available data on taxon ages.
<!--What led to datelife development?
Describe public and research necessities covered by datelife
It is important to use available data on time frame of lineage origin:
To know the state of dating for a group of interest:
What range of estimated ages exist already? Are fossil and molecular
time frames coherent? (e.g., [@Magallon2015c]).
To construct a time tree of life.
For science communication, improve scientific discussions, time-framing
other events of importance in other research areas.
Comparative method: autocorrelation (phylogenetic relatedeness) corrections, phylogenetic signal
-->

<!--INTRODUCTION: 4th paragraph-->
The TimeTree project [@Hedges2006; @Hedges2015; @Kumar2017] has aggregated chronograms
from 3,163 studies, encompassing 97,085 species [@Kumar2017], and continues to grow.
However, even in this gold standard resource, the included taxa only encompass between
0.097 and 3.236% of total species diversity (following taxonomic expert opinion
on the global, extant species numbers, which ranges from 3 to 100 million species
[Mayr2010; Moran2011]). One advantage of TimeTree is that it includes taxa from
across the tree of life, versus more specialized chronograms focusing on plants
[PHYLOMATIC], birds [JETZ ET AL BIRDTREE.ORG], and other groups. Users can choose
between a web interface or a mobile app to receive information on divergence times
for the evolutionary history of a lineage, pairs of taxa, all lineages within a
taxon, or a list of taxa. As a science communication tool, TimeTree project is
very powerful: it has a friendly graphical interface, with informative and colorful
outputs, that allows the general public to satisfy curiosity regarding a particular
organism of interest or group of them. It is of limited utility for scientific
studies, however. The thousands of trees that have been entered are unavailable
for examination or reuse; according to the creators (see TimeTree web FAQ), methods
for allowing data downloading have been under discussion for the past several years
yet the primary data remain closed. Moreover, there is no Application Programming
Interface (API) allowing programmatic access to any data, greatly impairing the
possibility of large-scale, automated data-mining, which is not allowed under TimeTree
website's terms of use. The nearly hundred thousand taxon summary chronogram generated
from TimeTree resources is not available with its publication [@Kumar2017] or the
TimeTree website, though the still substantial chronogram from a previous publication
[@Hedges2015] was made available at OToL.

With this inspiration, a prototype DateLife service was developed over a series
of phylotastic hackathons [__CITE__] at the National Evolutionary Synthesis Center.
NSF funding allowed for further development leading to this paper. A core goal of
DateLife is openness of both the data sources and the code underlying the analyses,
so scientific community can take advantage of this tool to leverage available information
to the advancement of discoveries in biology.

<!--
[@Mora2011]
eukaryotes and prokaryotes:
on earth: 1.24 million catalogued, 8.74 million predicted
on the oceans: 0.19 million catalogued, 2.21 predicted

Total ssp number estimated to be between 3 and 100 million species, as mentiones in [@Mora2011]:
ttol_spp <- 97085
(ttol_spp * 100) / 3e+6  # 3.236167
(ttol_spp * 100) / 100e+6  # 0.097085
-->
<!--
While its node-time search is very powerful, using available data to infer divergence
time from lineages that have never been sequenced before, its timetree option is
less powerful, their main flaw being that it does not provide source data.
However, their primary database is not fully open, giving information on a per node
basis. So, it is not possible to easily access information from original studies
for a group of taxa (One would need to perform a pair of taxa search for all pairs
from the group of interest).
Divergence time estimates for all taxa in the TToL are precomputed and stored in
the TimeTree database
All searches allow species names and higher group names. Common names are allowed
in all searches but a list of taxa.
one-taxon search was not working for dog, though.
1) ttol one lineage: it gives the evolutionary timeline of a lineage, showing the
age of main divergence events occuring along the lineage evolutionary history, i.e.,
when it shared a comon ancestor with all main lineages from which it diverged a
long time ago.
It only shows results graphically on the website.
2) ttol parwise search: allows common names, species names and higher group names
    how it handles names missing from tree: it looks in the NCBI taxonomy for the
    closest relative of the species or higher taxa. Then uses the closest relative
    to get the lineage pair age from ttol.
3) ttol group of taxa
4) ttol list of taxa: Cannot get it to work; does not work with cat, dog, turtle
search, felidae, hominidae, phrynops search, some_ants species name search. What
kind of files does it accepts: txt, rtf, doc, docx?
-->

<!--
retrieves information takes available data on lineage ages and phylogenetic relationships
to construct a tree of all life scaled to absolute time: a chronogram of all life,
while allowing the general public to search for age data within the tree for one
taxon of interest (timeline option), a pair of taxa (node time option), a group
of taxa (it can only be a single name of a taxonomic group), or a list of taxa of
interest (timetree option for the two latter).
Good things:
Allows to get ages of species with no data, by extrapolating from closely related
species with data.
Allows comparison of timelines to geologic events, such as asteroid impacts, oxygen
and carbon dioxide levels, solar luminosity.
Not so good things:
Only allows to download some data from original sources if performing a pair-of-taxa
search.
Only allows to see age ranges or confidence intervals in graphical form, when hovering
over nodes. These CIs come from ttol. It does not have CIs from original studies.
Node hovering is activated on one-taxon evolutionary history, and group-of-taxa
searches (we could not get to work list-of-taxa search). Pair-of-taxa search shows
the confidence interval of the only ttol node that constitutes the taxon-pair mrca.
You can only download data from pair-of-taxa and group-of-taxa search. It generates
a table with ages from original studies and a tree pruned from ttol, respectively.
In both cases it gives single point age estimates. We cannot download CI information
from ttol.
-->

Despite its great importance, analytical tools to summarize available information
on taxon ages for the scientific community are still lacking.
We identified several aspects that might have so far delayed the exploitation of
existing data. First, original chronograms available publicly are scattered across
various repositories (otol tree store, dryad, treebase, journals supplementary data)
usually with different formats too.
Second, lineage names due to taxonomic idiosincracy can be different among studies
and manual curation of that is usually necessary.
Third, data curation
Recent advances on this area (e.g., supersmart<!--and, which others?-->) aim to:
Generate new dates using all available DNA sequence information;
Perform one global analysis using all available information;
Problems or downsides: This might be time consuming for large groups and a lot of
data curation and knowledge on the group of interest is still necessary. For example,
<!--General issues with dating techniques:--> choosing correct fossils for calibration 
requires a lot of expertise and knowledge on the group. An incorrect use of fossils
can generate severe bias in dating results [@Sauquet2012c].
<!--Data curation is necessary at some point (is this always true?)-->
Hence, data curation is still an important part of any biological study. The research
community considers it as an important or even crucial step before data analysis.
Hence, automated processes for large data analysis are frequently received with skepticism.
<!--INTRODUCTION: 5th paragraph-->

DateLife palliates this by only using information available from already published
studies, which are ideally constructed using robust information, such as sequence
data and thoughtfully curated fossil calibrations.
<!--Data curation is largely based on taxonomic knowledge; or at least on the interplay between taxonomy and phylogenetics-->
<!--Importance of datelife:-->
DateLife can summarize this information in several formats that can be easily inspected
by users.
This allows rapidly obtaining a time frame of lineage divergence for a wide number
of taxa.
DateLife can also generate chronograms for taxa with little available information,
by using the available data as calibration points.
DateLife is the main service for scaling phylogenetic trees in Phylotastic! system
[@Stoltzfus2013] <!--to construct phylogenetic trees on the fly [http://phylotastic.org/].-->
It can be used through an R package <link to documentation>, a web interface
(<http://www.datelife.org/query/>) and an API<!--(still not up, right?)-->.
In here we present the first release of DateLife.
It contains an improved database of chronograms, more methods to summarize trees,
and new functions to visualize data.
It also allows comparison of summary methods.

<!--DESCRIPTION of Datelife-->
<span style="font-variant:small-caps;">
Description
</span>
```{r, echo = FALSE}
fig_workflow_main_foot <- paste("Stylized DateLife workflow. Detailed steps are shown in",  fig_workflow_all_functions)
```
<!--(BOLD and OToL species names, are they homogeneous/standardized?-->
The basic `DateLife` workflow is shown in `r fig_workflow_main` and consists of:

@@

<!--To perform the former tasks:-->
`DateLife`'s chronogram database is currently built from Open Tree of Life (OToL)'s [@Hinchliff2015] tree repository. Among currently existing repositories (e.g., TreeBAse, Dryad), OToL's metadata rich tree store  is the only one meeting the requirements for proper/accurate automatized handling of trees.
Input taxon names accepted by `DateLife` are binomial species names or clades. <!--common names (implemented yet?)--> Taxon searches are performed at the species level, so when input names correspond to higher clades, `DateLife` pulls all accepted species names within the
clade from OToL's reference taxonomy to perform the search.
Input names belonging to subspecies or any other infraspecific category are treated as
species.
DateLife also processes input names with the taxon name resolution service (TNRS),
<!--which "corrects misspelled names and authorities, standardizes variant spellings,
and converts nomenclatural synonyms to accepted names" (Boyle et al. 2013)-->
which corrects potentially misspelled names and typos, and standardizes variation in spelling
and synonyms <!--to a taxonomic reference of users choosing, degfault to otol taxonomy -->[@Boyle2013], increasing the probability to correctly find the queried
taxa in `DateLife`'s chronogram database.

Source chronogram summary tree can be assembled using the Super Distance Matrix (SDM) supertree construction approach [@Criscuolo2006] or using the median of branch lengths and the hierarchical clustering method <!-- of ttol-->. <!--Also, I just discovered the median gives non ultrametric trees... but I thonk it is fixed now, check it!).--> 
Tree dating and simulation options are performed with various algorithms:
Branch Length Adjuster (BLADJ) is a simple algorithm to distribute ages of undated
nodes evenly, which minimizes age variance in the chronogram [@Webb2008].
PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees. 
treePL, is a semi-parametric, rate-smoothing, penalized likelihood dating method
[@Smith2012].
MrBayes [@Huelsenbeck2001; @Ronquist2003] can be used when adding taxa at
random, following a reference taxonomy or a topological constraint. It draws ages
from a pure birth model, as implemented by Jetz and collaborators [-@Jetz2012].
To apply calibrations to a tree, congruification method described in [@Eastman2013] to find shared nodes between trees (congruent nodes).

To gather, process, and present information, `DateLife` builds up from functions available in several R packages including rotl [@Michonneau2016], ape[@Paradis2004], geiger[@Harmon2008], paleotree[@Bapst2012a], bold[@Chamberlain2018], phytools[@Revell2012], taxize[@Chamberlain2013; @Chamberlain2018], phyloch[@Heibl2008], phylocomr[@Ooms2018] and rphylotastic[@Omeara2019].

@@

DateLife's code speed was tested on an Apple iMac
with one 3.4 GHz Intel Core i5 processor. <!--No multiprocessor tests were performed-->
We registered variation in computing time relative to number of input names
<!--number of chronograms in database--> and DateLife service.
Input processing increases roughly linearly with number of input taxon names, and
increases considerably if tnrs service is activated (`r fig_runtime_main`).
<!--
Speed with different number of input lineages and types of analysis:
-->
Results show that searching time increases linearly with number of input names and
number of chronograms in database.

@@

<!--
Speed of web interface and of r package (in computers with different capacities?)
-->
DateLife's code performance was evaluated with a set of unit tests designed and
implemented with the R package testthat [@RCoreTeam2018]. These tests were run both
locally –using the devtools package [@RCoreTeam2018]– and on a public server –via
GitHub– using the continuous integration tool Travis CI (<//travis-ci.org>). At
present, unit tests cover around 30% (for now) of DateLife's code (<https://codecov.io/gh/phylotastic/datelife>). <!-- unit test make code robust, verifiable, debuggable, -->

@@

```{r pressure, out.width="100%", include=TRUE, fig.align="center", fig.cap=fig_workflow_main_foot, echo=FALSE}
knitr::include_graphics("fig1.pdf", dpi = 360)
```

>>>>

---
output: pdf_document
fontsize: 12pts
textalign: left  # this does not work...
geometry: margin = 1in
bibliography: library_red.bib
csl: systematic-biology.csl
link-citations: yes
# indent: true # this works
# header-includes:
#    - \setlength\parindent{24pt}  # this works too, but I'm gonna indent manually with 4 spaces
# nope, 4 spaces indents are not recognized, they do something else
linestretch: 2
header-includes:
    - \usepackage[left]{lineno}
    - \linenumbers
---
```{r eval = TRUE, echo = FALSE}
source("Datelife_paper1_global.R")
```

@@

<!--INTRODUCTION-->
Supersmart is an open tool but not easy to use, still requires a lot of curation and knowledge.

<!--INTRODUCTION: 1st paragraph: general statement that the field knows to be true and important-->
Time of lineage divergence<!--alt lineages: named biological entities, taxon ages?-->
constitutes in many ways the fundamental/main knowledge necessary for evolutionary
understanding.
<!--Along with phylogenetic relationships, it is the most important piece of information
allowing to test alternative evolutionary hypothesis in many areas of research,
from developmental to ecological questions.
Along with migration rates and patterns, it is fundamental to test biogeographical
hypothesis-->
Coupled to species number and distribution, it is the main information necessary
for the study of diversification processes (i.e., the tempo and mode of speciation
and extinction),
<!-- crucial/vital/critical/essential  -->central for the understanding of how biodiversity
patterns are shaped across space, time and clades [@Morlon2014].
Evolutionary understanding also relies on comparative studies, for which knowing
the time context for all life is crucial. Efforts to have a whole tree of life have
been great and here are some examples.
<!--INTRODUCTION: 2nd paragraph: important advances in the field-->
In the past two decades, the possibility to obtain good quality DNA sequences
<!-- from a wide variety of organisms became a reality, which,  -->
coupled to methodological developments in phylogenetic and dating inference, allowed
the application of molecular dating methods on a very large amount and diversity
of organisms, greatly increasing the quantity of<!--organisms with--> data on taxon
ages across the tree of life.
<!-- figure showing growth on studies with age data from references table from ttol-->
To date, there is a large amount of both fossil and molecular-based data on taxon
ages and phylogenetic relationships in public repositories such as Dryad, TreeBASE
and Open Tree of Life (OToL).
OToL alone holds more than 200 chronograms. <!--cache updated jan 15 2018
devtools::load_all(pkg = "~/Desktop/datelife")
length(opentree_chronograms$trees) names(opentree_chronograms)
-->
<!--How many trees? How many dated trees?-->
Methods to include living and fossil lineages are in continued development and increased
usage by the community<!--total evidence and revbayes (fossilized birth-death)-->,
which coupled to better sharing data practices, are greatly contributing to the
accumulation in number and type of available data on taxon ages.

<!--INTRODUCTION: 3rd paragraph: problems associated-->
The TimeTree project [@Hedges2006; @Hedges2015; @Kumar2017] has aggregated chronograms
from 3,163 studies, encompassing 97,085 species [@Kumar2017], and continues to grow.
However, even in this gold standard resource, the included taxa only encompass between
0.097 and 3.236% of total species diversity (following taxonomic expert opinion
on the global, extant species numbers, which ranges from 3 to 100 million species
[Mayr2010; Moran2011]). One advantage of TimeTree is that it includes taxa from
across the tree of life, versus more specialized chronograms focusing on plants
[PHYLOMATIC], birds [JETZ ET AL BIRDTREE.ORG], and other groups. Users can choose
between a web interface or a mobile app to receive information on divergence times
for the evolutionary history of a lineage, pairs of taxa, all lineages within a
taxon, or a list of taxa. As a science communication tool, TimeTree project is
very powerful: it has a friendly graphical interface, with informative and colorful
outputs, that allows the general public to satisfy curiosity regarding a particular
organism of interest or group of them. It is of limited utility for scientific
studies, however. The thousands of trees that have been entered are unavailable
for examination or reuse; according to the creators (see TimeTree web FAQ), methods
for allowing data downloading have been under discussion for the past several years
yet the primary data remain closed. Moreover, there is no Application Programming
Interface (API) allowing programmatic access to any data, greatly impairing the
possibility of large-scale, automated data-mining, which is not allowed under TimeTree
website's terms of use. The nearly hundred thousand taxon summary chronogram generated
from TimeTree resources is not available with its publication [@Kumar2017] or the
TimeTree website, though the still substantial chronogram from a previous publication
[@Hedges2015] was made available at OToL.
<!--What led to datelife development?
Describe public and research necessities covered by datelife
It is important to use available data on time frame of lineage origin:
To know the state of dating for a group of interest:
What range of estimated ages exist already? Are fossil and molecular
time frames coherent? (e.g., [@Magallon2015c]).
To construct a time tree of life.
For science communication, improve scientific discussions, time-framing
other events of importance in other research areas.
Comparative method: autocorrelation (phylogenetic relatedeness) corrections, phylogenetic signal
-->

<!--INTRODUCTION: 4th paragraph-->
Despite its great importance, analytical tools to summarize available information
on taxon ages for the scientific community are still lacking.
We identified several aspects that might have so far delayed the exploitation of
existing data. First, original chronograms available publicly are scattered across
various repositories (otol tree store, dryad, treebase, journals supplementary data)
usually with different formats too.
Second, lineage names due to taxonomic idiosincracy can be different among studies
and manual curation of that is usually necessary.
Third, data curation
Recent advances on this area (e.g., supersmart<!--and, which others?-->) aim to:
Generate new dates using all available DNA sequence information;
Perform one global analysis using all available information;
Problems or downsides: This might be time consuming for large groups and a lot of
data curation and knowledge on the group of interest is still necessary. For example,
<!--General issues with dating techniques:--> choosing correct fossils for calibration
requires a lot of expertise and knowledge on the group. An incorrect use of fossils
can generate severe bias in dating results [@Sauquet2012c].
<!--Data curation is necessary at some point (is this always true?)-->
Hence, data curation is still an important part of any biological study. The research
community considers it as an important or even crucial step before data analysis.
Hence, automated processes for large data analysis are frequently received with skepticism.

DateLife palliates this by only using information available from already published
studies, which are ideally constructed using robust information, such as sequence
data and thoughtfully curated fossil calibrations.
<!--Data curation is largely based on taxonomic knowledge; or at least on the interplay between taxonomy and phylogenetics-->
<!--Importance of datelife:-->

<!--INTRODUCTION: 5th paragraph: Goal, objectives and predictions-->
Rapidly increasing data on time of lineage divergence both from molecular and paleontological studies; the increasing importance of use of these data in distant areas of research, often not specialized enough to rapidly obain data on their own; and the lack of an open (both the data sources and the code underlying the analyses) easy to use tool
inspired the development of a prototype `DateLife` service over a series
of phylotastic hackathons [@Stoltzfus2013] at the National Evolutionary Synthesis Center.
In this paper we present the first formal description of `DateLife`, featuring an improved database of chronograms, more methods to summarize trees, and new functions to visualize data, as well as comparisons of summary trees.
DateLife is the main service for scaling phylogenetic trees in Phylotastic! system
[@Stoltzfus2013] <!--to construct phylogenetic trees on the fly [http://phylotastic.org/].-->
It can be used through an R package <link to documentation>, a web interface
(<http://www.datelife.org/query/>) and an API<!--(still not up, right?)-->.

<!--DESCRIPTION of Datelife-->
<span style="font-variant:small-caps;">
Description
</span>

<!--(BOLD and OToL species names, are they homogeneous/standardized?-->
The basic `DateLife` workflow is shown in `r fig_workflow_main` and consists of:

@@

<!--To perform the former tasks:-->
`DateLife`'s chronogram database is currently built from Open Tree of Life (OToL)'s
[@Hinchliff2015] tree repository. Among currently existing repositories (e.g., TreeBAse,
Dryad), OToL's metadata rich tree store is the only one meeting the requirements
for proper/accurate automatized handling of trees.
Input taxon names accepted by `DateLife` are binomial species names or clades.
<!--common names (implemented yet?)-->
Taxon searches are performed at the species level, so when input names correspond
to higher clades,
`DateLife` pulls all accepted species names within the
clade from OToL's reference taxonomy to perform the search.
Currently, searches at the infrapsecies level are not allowed, so input names belonging to subspecies or any other infraspecific category are treated
as species.
`DateLife` also processes input names with the taxon name resolution service (TNRS),
<!--which "corrects misspelled names and authorities, standardizes variant spellings,
and converts nomenclatural synonyms to accepted names" (Boyle et al. 2013)-->
which corrects potentially misspelled names and typos, and standardizes variation
in spelling and synonyms <!--to a taxonomic reference of users choosing, default
to otol taxonomy -->[@Boyle2013], increasing the probability to correctly find the
queried
taxa in `DateLife`'s chronogram database.

Source chronogram summary tree can be assembled using the Super Distance Matrix
(SDM) supertree construction approach [@Criscuolo2006] or using the median of branch
lengths and the hierarchical clustering method<!-- of ttol-->. <!--Also, I discovered the median gives non ultrametric trees... but I thonk it is fixed now, check it!).-->
Tree dating and simulation options are performed with various algorithms:
Branch Length Adjuster (BLADJ) is a simple algorithm to distribute ages of undated
nodes evenly, which minimizes age variance in the chronogram [@Webb2008].
PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees.
treePL, is a semi-parametric, rate-smoothing, penalized likelihood dating method
[@Smith2012].
MrBayes [@Huelsenbeck2001; @Ronquist2003] can be used when adding taxa at
random, following a reference taxonomy or a topological constraint. It draws ages
from a pure birth model, as implemented by Jetz and collaborators [-@Jetz2012].
To apply calibrations to a tree, th econgruification algorithm described in [@Eastman2013]
is used to find shared nodes between trees (congruent nodes).

To gather, process, and present information, `DateLife` builds up from functions
available in several R packages including rotl [@Michonneau2016], ape [@Paradis2004],
geiger [@Harmon2008], paleotree [@Bapst2012a], bold [@Chamberlain2018], phytools [@Revell2012],
taxize [@Chamberlain2013; @Chamberlain2018], phyloch [@Heibl2008], phylocomr [@Ooms2018]
and rphylotastic [@Omeara2019].

@@

`DateLife`'s code speed was tested on an Apple iMac
with one 3.4 GHz Intel Core i5 processor. <!--No multiprocessor tests were performed-->
We registered variation in computing time relative to number of input names
<!--number of chronograms in database--> and `DateLife` service.
Input processing increases roughly linearly with number of input taxon names, and
increases considerably if tnrs service is activated (`r fig_runtime_main`).
<!--
Speed with different number of input lineages and types of analysis:
-->
Results show that searching time increases linearly with number of input names and
number of chronograms in database.

@@

<!--
Speed of web interface and of r package (in computers with different capacities?)
-->
`DateLife`'s code performance was evaluated with a set of unit tests designed and
implemented with the R package testthat [@RCoreTeam2018]. These tests were run both
locally –using the devtools package [@RCoreTeam2018]– and on a public server –via
GitHub– using the continuous integration tool Travis CI (<//travis-ci.org>). At
present, unit tests cover around 30% (for now) of `DateLife`'s code (<https://codecov.io/gh/phylotastic/datelife>). <!-- unit test make code robust, verifiable, debuggable, -->

@@
